---
title: "Theory-based Quantitative Data"
author: "Mark Schulist"
date: "`r Sys.Date()`"
output: rmdformats::downcute
downcute_theme: "chaos"
---

We got the number of minutes that it takes for us to get to school. 

```{r}
minutes <- c(9,18,15,23,5,9,15,5,45,27,70,17,28,8,22,16,25,16)
```

Now we'll get the bootstrap mean distribution.

Empirical Method (Simulated):
```{r}
means <- replicate(5000, mean(sample(minutes, replace = T)))
upper_emp <- quantile(means, c(0.025, 0.975))[2]
lower_emp <- quantile(means, c(0.025, 0.975))[1]

print(paste0("empirical: ", "(", round(lower_emp, 4), ", ", round(upper_emp, 4), ")"))
```

Theoretical method:
```{r}
upper <- mean(minutes) + 2 * sd(means)
lower <- mean(minutes) - 2 * sd(means)
print(paste0("Theoritical: ", "(", round(lower, 4), ", ", round(upper, 4), ")"))
```

# Using t-distribution

This is the formula for approximating the sd of the null distribution. 
$$
\frac{s}{\sqrt{n}}
$$

```{r}
sd_theory <- sd(minutes) / sqrt(length(minutes))
sd_theory
```

This compares well to the sd of the means (`r round(sd(means), 4)`). 


Conducting a t-test.
```{r}
t_test <- t.test(minutes, alternative = "greater", mu = 30)
t_test
```

So our p-value is `r t_test$p.value`. 

We can calculate the t-stat (which is the same thing as the z-score for the t-distribution). This value is also in the above code as the "t". 

```{r}
(mean(minutes) - 30) / sd_theory
```

We will now graph a box plot of our sample data to see how it relates to our t stats. 

```{r}
boxplot(minutes, horizontal = T)
```

# Conclusion

We need to have sample size of 20 or more or a symmetric sample distribution to use the t-test. 