---
title: "Chi-squared Tests"
author: 
date: 
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Goodness of fit test

To perform a goodness of fit test in R, we need to provide a vector containing the observed counts and a vector containing the hypothesized distribution.  For our jury example:

```{r}
obs_counts <- c(205, 26, 25, 19)
probs <- c(0.72, 0.07, 0.12, 0.09)
```

The hypothesis test is performed using the `chisq.test()` function.

```{r}
chisq.test(obs_counts, p = probs)
```

We can obtain the expected counts in R by performing the hypothesis test and assigning the results to an object.

```{r}
t <- chisq.test(obs_counts, p = probs)
t
```

This object contains many items, two of which are the p-value and the expected counts.

```{r}
t$p.value
t$expected
```

```{r}
bites <- c(24750, 25250, 24800, 25200)
moon_probs <- rep(1/4, 4)
sum(bites)
bites/sum(bites)
```

```{r}
d <- chisq.test(bites, p = moon_probs)
d
```



## Tests of Independence/Dependence

To perform a test of independence/dependence, we need to provide R with a matrix containing our observed counts.  For the depression example:

```{r}
obs_counts <- c(16, 29, 9, 22, 33, 14, 19, 14, 3)
obs_counts
```

When creating `obs_counts`, enter the numbers by columns, not by row.

```{r}
obs_matrix <- matrix(obs_counts, nrow = 3, ncol = 3)
obs_matrix
```

The hypothesis test is also performed using the `chisq.test()` function.

```{r}
chisq.test(obs_matrix)
```

## Immigration Example

We can load in the dataset `immigration`:

```{r}
source("https://www.openintro.org/data/R/immigration.R")
```

Let's start by briefly exploring the dataset:

```{r}
head(immigration)
```

```{r}
summary(immigration)
```

We can obtain the contingency table that cross-classifies respondents using `table()`.

```{r}
table(immigration$response, immigration$political)
table(immigration)
```

If we assign our table a name, we can send it directly to the `chisq.test()` function.

```{r}
obs_table <- table(immigration)
t <- chisq.test(obs_table)
t
```

Let's check our expected counts:

```{r}
t$expected
```

We can remove the `Not sure` category from our analysis, and redo our hypothesis test.

```{r}
obs_table2 <- table(immigration, exclude = "Not sure")
obs_table2
```

```{r}
t2 <- chisq.test(obs_table2)
t2
t2$expected
```

## Example 1

```{r}
obs_count <- c(17, 26, 22, 23, 19, 15, 25)
probs <- rep(1 / length(obs_count), length(obs_count))
chisq.test(obs_count, p = probs)
```

## Example 2

```{r}
obs_count <- matrix(c(113,92,934,1070), ncol = 2, nrow = 2)
t <- chisq.test(obs_count)
t
```



## Comparing Proportions

```{r}
left_handed <- c(113, 92)
right_handed <- c(1047, 1162)
prop.test(left_handed, right_handed, alternative = "greater")
```

Creating our matrix of observed counts:

```{r}
obs_counts <- c(370, 218, 480, 362)
obs_matrix <- matrix(obs_counts, nrow = 2, ncol = 2)
obs_matrix
```

Doing our chi-squared test:

```{r}
chisq.test(obs_matrix)
```

