---
title: "Homework 6"
author: "Mark Schulist"
format: pdf
---

1. 

```{r}
source("https://www.openintro.org/data/R/speed_gender_height.R")
```

```{r}
boxplot(speed_gender_height$speed ~ speed_gender_height$gender)
```

The boxplots appear mostly normal, especially given the large sample sizes for each group. There is some left skewedness, but it is not too bad. 

$$
H_0: \mu_1 = \mu_2
$$
$$
H_a: \mu_1 \neq \mu_2
$$

```{r}
t.test(speed_gender_height$speed ~ speed_gender_height$gender)
```

The p-value is extremely small, so we reject the null hypothesis and have evidence to show that there is a true different in means between male and female drivers' top speed driven. 

$$
H_0: \tilde{\mu}_1 = \tilde{\mu}_2
$$
$$
H_a: \tilde{\mu}_1 \neq \tilde{\mu}_2
$$

```{r}
wilcox.test(speed_gender_height$speed ~ speed_gender_height$gender)
```

The p-value is very small, so so we reject the null hypothesis and have evidence to show that there is a true different in medians between male and female drivers' top speed driven. 

2. 

```{r}
source("https://www.openintro.org/data/R/hsb2.R")
```

```{r}
boxplot(hsb2$read ~ hsb2$schtyp)
```

The boxplots appear mostly normal, although there is some right skew in both of them. Given we have 100 samples for each group, this should be fine. 

$$
H_0: \mu_1 = \mu_2
$$
$$
H_a: \mu_1 \neq \mu_2
$$

```{r}
t.test(hsb2$read ~ hsb2$schtyp)
```

The p-value is above 0.05, so we do not reject the null hypothesis. We do not have evidence to conclude that the true difference in mean reading scores between public and private schools is different. 

$$
H_0: \tilde{\mu}_1 = \tilde{\mu}_2
$$
$$
H_a: \tilde{\mu}_1 \neq \tilde{\mu}_2
$$


```{r}
wilcox.test(hsb2$read ~ hsb2$schtyp)
```

The p-value is above 0.05, so we do not reject the null hypothesis. We do not have evidence to conclude that the true difference in median reading scores between public and private schools is different. 

3. 

```{r}
boxplot(hsb2$read, 
        hsb2$write, 
        names = c("read", "write"), 
        main = "Reading and Writing Scores"
        )
```

The writing median appears to be higher, but overall the distributions look very similar. The writing scores are left skewed and the reading scores are (a little) right skewed. 

Our samples are not independent---they are paired. 

Because the reading and writing distributions are (mostly) normal, the assumption of normality is reasonable. There are also 100 samples, which is plenty. 

$$
H_0: \mu_{\mathrm{diff}} = 0
$$
$$
H_a: \mu_{\mathrm{diff}} \neq 0
$$

```{r}
t.test(hsb2$read, hsb2$write, paired = T)
```

The p-value is above 0.05, so we do not reject the null. We do not have evidence to suggest that the difference between each person's reading and writing scores is not equal to 0. 

4. 

```{r}
source("https://www.openintro.org/data/R/climate70.R")
```


```{r}
boxplot(climate70$dx90_1948, climate70$dx90_2018, names = c("1948", "2018"))
```

There is not a large difference between the distributions, although 2018 seems to have a longer right tail. These distributions are not normal, although we have enough samples (197) that the normality is probably not a big concern.  

We have paired samples. They are from the same locations on different dates. 

$$
H_0: \mu_{\mathrm{diff}} = 0
$$
$$
H_a: \mu_{\mathrm{diff}} \neq 0
$$

```{r}
t.test(climate70$dx90_1948, climate70$dx90_2018, paired = T)
```

The p-value is below $\alpha$, so we reject the null hypothesis. We have evidence that there is a difference in the average number of days above 90 degrees between 1948 and 2018. 

5. 

a. Paired, they come from the same days. 

b. Paired, they come from the same items. 

c. Not paired, the students are independent in both samples. 

d. Not paired, the salaries are independent in both samples. 

e. Paired, the people are the same in both samples, just at different points in time. 
