---
title: "Hypothesis Tests"
author: 
date: 
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Lupine Example Hypothesis Test

For this example we had 142 out of 200 plants that were blue flowering.  Our hypothesized value was 0.75.

```{r}
prop.test(142, 200, p = 0.75, correct = FALSE)
```

R does not report Z, instead it reports Z^2, which it calls X-squared.

## Nuclear Arms Reduction

```{r}
prop.test(576, 1028, p = 0.5, correct = FALSE, alternative = "greater")
```

If our alternative hypothesis is p < p0, we would use the `alternative = "less"` option.

## Uniformity of p-values

Suppose in our population the true value of p is 0.3.  We take a random sample of 5000 individuals in our population and perform a hypothesis test:
$$H_0: p = 0.3 \\ H_a: p \ne 0.3$$
Note here that H0 is true.

```{r}
# sample size
n <- 5000

# generate the number of successes for our random sample
X <- rbinom(1, n, 0.3)
```

```{r}
# perform our hypothesis test
test <- prop.test(X, n, p = 0.3, correct = "FALSE")
test
```

We can obtain just the p-value from our hypothesis test:

```{r}
test$p.value
```

Now, let's take 100,000 random samples of size 5000, perform 100,000 hypothesis tests, and obtain 100,000 p-values.

```{r}
n <- 5000
samples <- 100000

set.seed(6)
X <- rbinom(samples, n, 0.3)
```

X contains the number of successes in each of our 100,000 random samples.

```{r}
# a vectors for our p-values to go in
pvalues <- rep(0, samples)    

# performing 100,000 tests and getting 100,000 p-values
for(i in 1:samples){
  test <- prop.test(X[i], n, p = 0.3, correct = "FALSE")
  pvalues[i] <- test$p.value
}
```

Now, let's make a histogram of the p-values:

```{r}
hist(pvalues)
```

## Small Sample Size

We can use `binom.test()` to perform a hypothesis test for the organ donation example.

```{r}
binom.test(3, 62, p = 0.10, alternative = "less")
```


Now, let's take 100,000 random samples of size 10, perform 100,000 hypothesis tests, and obtain 100,000 p-values.  We will still assume that in our population the true value of p is 0.3, and we will still test the hypotheses:
$$H_0: p = 0.3\\ H_a: p \ne 0.3$$

```{r}
n <- 10
samples <- 100000

set.seed(341)
X <- rbinom(samples, n, 0.3)
```

X contains the number of successes in each of our 100,000 random samples.

```{r}
# a vectors for our p-values to go in
pvalues <- rep(0, samples)    

# performing 100,000 tests and getting 100,000 p-values
for(i in 1:samples){
  test <- suppressWarnings(prop.test(X[i], n, p = 0.3, correct = "FALSE"))
  pvalues[i] <- test$p.value
}
```

Now, let's make a histogram of the p-values:

```{r}
hist(pvalues)
```

## Power

R has some built in functions to help us compute power.  The function we will use for a hypothesis test for a single proportion is in the package `pwr`.  We first need to install this package using the command `install.packages("pwr")`.  Installing a package only needs to be done once.  However, we will need to load this package every time we use it.

```{r}
library(pwr)
```

The function we will use is `pwr.p.test()`.  The syntax:

```{r}
# 0.65 is the true value of the proportion
# 0.6 is our hypothesized proportion
# 100 is our proposed sample size 
# 0.05 is our significance level
# We are doing a two.sided test (the other options are "less" and "greater")

pwr.p.test(ES.h(0.65, 0.60), n = 100, sig.level = 0.05, alternative = "two.sided")
```

Now let's modify some of our settings and see what happens.

```{r}
pwr.p.test(ES.h(0.65, 0.6), n = 1000, sig.level = 0.05, alternative = "two.sided")
```

We can also use `pwr.p.test()` to compute the sample size needed to reach a desired power.

```{r}
pwr.p.test(ES.h(0.65, 0.6), power = 0.90, sig.level = 0.05, alternative = "two.sided")
```

```{r}
pwr.p.test(ES.h(0.57, 0.6), sig.level = 0.05, n = 600, alternative = "less")
```

```{r}
pwr.p.test(ES.h(0.57, 0.6), sig.level = 0.05, power = 0.9, alternative = "less")
```



