---
title: "Probability"
author: 
date: 
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

The goals for this lab are to 

* learn how to simulate the outcomes of random processes in `R`

* think about the effects of independent and dependent events

* to compare a simulation to actual data

## Simulation

In a simulation, you set the ground rules of a random process and then the computer uses random numbers to generate an outcome that adheres to those rules. As a simple example, you can simulate rolling a fair die.

```{r}
outcomes <- c(1, 2, 3, 4, 5, 6)
fairdie_prob <- c(1/6, 1/6, 1/6, 1/6, 1/6, 1/6)
sample(outcomes, size = 1, replace = TRUE, prob = fairdie_prob)
```

The vector `outcomes` can be thought of as a hat with 6 slips of paper in it: one slip for each outcome of our die. The vector `fairdie_prob` contains the probabilities for each outcome. The function `sample` draws one slip from the hat and tells us if it was a 1, 2, 3, 4, 5, or 6.

Run the third command listed above several times. Just like when rolling a die, the number rolled will vary, but in the long run, you’d expect to get roughly equal numbers of each.

If you wanted to simulate rolling a fair die 100 times, you could either run the function `sample` 100 times or, more simply, adjust the `size` argument, which governs how many samples to draw (the replace = TRUE argument indicates we put the slip of paper back in the hat before drawing again). Save the resulting vector of rolls in a new object called `sim_fair_die`.

```{r}
set.seed(2211)
sim_fair_die <- sample(outcomes, size = 100, replace = TRUE, prob = fairdie_prob)
```

To view the results of this simulation, type the name of the object and then use table to count up the number of each outcome.

```{r}
sim_fair_die
table(sim_fair_die)
```

We can also simulate the outcomes from an unfair die.  Suppose we have a weighted die that rolls a 1 with probability 0.5, and rolls 2 through 6 each with probability 0.1.

```{r}
unfairdie_prob <- c(0.5, 0.1, 0.1, 0.1, 0.1, 0.1)
set.seed(23)
sim_unfair_die <- sample(outcomes, size = 10000, replace = TRUE, prob = unfairdie_prob)
```

### Exercise 1

Look at the results of the simulation for the unfair die, and use the `table` command to count the number of 1's that resulted.

```{r}
table(sim_unfair_die)
hist(sim_unfair_die,breaks = 0:6)
```


Recall: Multiple rolls of a die are independent random processes.

## Hot Hands

Basketball players who make several baskets in succession are described as having a hot hand. Fans and players have long believed in the hot hand phenomenon, which refutes the assumption that each shot is independent of the next. However, a 1985 paper by Gilovich, Vallone, and Tversky collected evidence that contradicted this belief and showed that successive shots are independent events (http://www.cs.colorado.edu/~mozer/Teaching/syllabi/7782/readings/gilovich%20vallone%20tversky.pdf). This paper started a great controversy that continues to this day, as you can see by Googling hot hand basketball.

We do not expect to resolve this controversy today. However, we’ll apply one approach to answering questions like this. 

Our investigation will focus on the performance of one player: Kobe Bryant of the Los Angeles Lakers. His performance against the Orlando Magic in the 2009 NBA finals earned him the title Most Valuable Player and many spectators commented on how he appeared to show a hot hand. Let’s load some data from those games and look at the first several rows. 

```{r}
download.file("http://www.openintro.org/stat/data/kobe.RData", destfile = "kobe.RData")
load("kobe.RData")
```

```{r}
kobe=kobe[order(kobe$game),]
head(kobe)
```


In this data frame, every row records one of the 133 shots taken by Kobe Bryant. If he hit the shot (made a basket), a hit, `H`, is recorded in the column named `basket`, otherwise a miss, `M`, is recorded.

Just looking at the string of hits and misses, it can be difficult to gauge whether or not it seems like Kobe was shooting with a hot hand. One way we can approach this is by considering the belief that hot hand shooters tend to go on shooting streaks. For this lab, we define the length of a shooting streak to be the number of consecutive baskets made until a miss occurs. A streak of length zero means a `M` with no `H` preceding it.

The custom function `calc_streak`, which was loaded in with the data, may be used to calculate the lengths of all shooting streaks and then look at the distribution.

```{r}
kobe_streak <- calc_streak(kobe$basket)
barplot(table(kobe_streak))
```

### Exercise 2

Describe the distribution of Kobe’s streak lengths from the 2009 NBA finals. What was his typical streak length? How long was his longest streak of baskets?

## Compared to What?

We’ve shown that Kobe had some long shooting streaks, but are they long enough to support the belief that he had hot hands? What can we compare them to?

To answer these questions, let’s return to the idea of independence. Two processes are independent if the outcome of one process doesn’t effect the outcome of the second. 

* If each shot that a player takes is an independent process, having made or missed your first shot will not affect the probability that you will make or miss your second shot.

* A shooter with a hot hand will have shots that are not independent of one another. Specifically, if the shooter makes his first shot, the hot hand model says he will have a higher probability of making his second shot.  As a result of these increased probabilities, you’d expect Kobe to have longer streaks. 

Now that we’ve phrased the situation in terms of independent shots, let’s return to the question: how do we tell if Kobe’s shooting streaks are long enough to indicate that he has hot hands? We can compare his streak lengths to someone without hot hands: an independent shooter.

## Simulating an Independent Shooter

We can simulate hits and misses from an independent shooter in the same way that we simulated rolls of a fair and unfair die.  

During his career, the percentage of time Kobe made a basket (i.e. his shooting percentage) was about 45%.  We can simulate his 133 shots assuming independent shots using `R`.

```{r}
outcomes <- c("H", "M")
basket_prob <- c(0.45, 0.55)
set.seed(10)
sim_basket <- sample(outcomes, size = 133, replace = TRUE, prob = basket_prob)
```

We can use the function `calc_streak` to calculate the lengths of the shooting streaks in our simulated data.  We can use `barplot` to look at the distribution.

```{r}
sim_streak <- calc_streak(sim_basket)
barplot(table(sim_streak))
```

### Exercise 3

How does Kobe Bryant’s distribution of streak lengths compare to the distribution of streak lengths for the simulated shooter? Using this comparison, do you have evidence that the hot hand model fits Kobe’s shooting patterns? Explain.

