---
title: "Multiple Regression and PCA"
author: 
date: 
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Births in 2014: Multiple Regression

We can load the dataset into R:

```{r}
source("https://www.openintro.org/data/R/births14.R")
```

We can make scatterplots for the response variable with the numeric predictors.

```{r}
plot(births14$fage, births14$weight)
plot(births14$mage, births14$weight)
plot(births14$weeks, births14$weight)
plot(births14$visits, births14$weight)
plot(births14$gained, births14$weight)
```

We can also create a *scatterplot matrix* for the numeric variables.

```{r}
num <- c(1, 2, 4, 6, 7, 8) # the columns 1, 2, 4, 6, 7, and 8 contain the numeric variables
pairs(births14[num])
```

We can create boxplots to look at the relationship between `weight` and the categorical predictors.

```{r}
boxplot(births14$weight ~ births14$premie)
boxplot(births14$weight ~ births14$sex)
boxplot(births14$weight ~ births14$habit)
boxplot(births14$weight ~ births14$marital)
boxplot(births14$weight ~ births14$whitemom)
```

Using `habit` as an explanatory variable:

```{r}
model1 <- lm(weight ~ habit, data = births14)
summary(model1)
```

```{r}
t.test(births14$weight ~ births14$habit, var.equal = T)
```


Estimating the multiple linear regression model:

```{r}
model2 <- lm(weight ~ fage + mage + weeks + premie + visits + gained + sex + habit + marital + whitemom, data = births14)
summary(model2)
```

Performing model selection:

```{r}
m <- lm(weight ~ mage + weeks + premie + gained + sex + habit + whitemom, data = births14)
summary(m)
plot(m)
```

```{r}
new <- data.frame(mage = 30, weeks = 36, premie = "full term", gained = 20, sex = "male", habit = "nonsmoker", whitemom = "not white")
predict(m, newdata = new, interval = "prediction")
```


## USArrests: Principal Component Analysis

```{r}
View(USArrests)
```

In this case, the number of variables is fairly small, so we can look at a scatterplot matrix.

```{r}
pairs(USArrests)
```

We can obtain the sample variances for our four variables.

```{r}
var(USArrests$Murder)
var(USArrests$Assault)
var(USArrests$UrbanPop)
var(USArrests$Rape)
```

The `prcomp` performs PCA.  The option `scale = TRUE` scales all the variables to have a standard deviation of 1.

```{r}
pca_result <- prcomp(USArrests, scale = TRUE)
pca_result
```

Plotting the explained variances:

```{r}
plot(pca_result, type = "l")
```

Making a biplot:

```{r}
biplot(pca_result, scale = 0)
```


 
 
 
 
 
