---
title: "Homework 2"
author: "Mark Schulist"
date: "`r Sys.Date()`"
format: typst
pdf-engine: typst
---


## 1.6.16

```{r}
salaries <- c(
  152, 169, 178, 179, 185, 188, 195, 196, 198, 203, 204, 209, 210, 212, 214
  )
```

a. 

```{r}
xbar <- mean(salaries)
s2 <- var(salaries)
```

The mean is `r xbar` and the variance is `r s2`.

b. 

```{r}
salaries_5000 <- salaries + 5
salaries_5 <- salaries * 1.05
```

i. The mean is `r mean(salaries_5000)` and the variance is `r var(salaries_5000)`.

ii. The mean is `r mean(salaries_5)` and the variance is `r var(salaries_5)`.


## 1.7.4

```{r}
si <- read.table(
  "https://media.pearsoncmg.com/cmg/pmmg_mml_shared/mathstatsresources/Akritas/SolarIntensAuData.txt",
  header = T
  )
```

a. 

```{r}
boxplot(si$SI, main="Solar Intensity Measurements")
```

b.

```{r}
quants <- quantile(si$SI, c(0.3, 0.6, 0.9))
```

The 30th quantile is `r quants[1]`, the 60th `r quants[2]`, and the 90th `r quants[3]`.

## 2.2.10 

a. 

- $E_1 = 21$
- $E_2 = 14$
- $E_3 = 30$
- $RR$

c. 

- $E_1 \cap E_2$: Disks that are both low hardness and low shock absorption. 
- $E_1 \cup E_2$: Disks that are low hardness or low shock absorption. 
- $E_1 - E_2$: Disks that are low hardness but not low shock absorption. 
- $(E_1 - E_2)  (E_2 - E_1)$: Disks that are either low hardness or low shock absorption, but not both. 

d. 

- $E_1 union E_2$: 5
- $E_1 \cup E_2$: 30
- $E_1 - E_2$: 16
- $(E_1 - E_2) \cup (E_2 - E_1)$: 25

## 2.3.9

a. ${12 \choose 4} = 495$

b. ${5 \choose 2} \cdot {4 \choose 1} \cdot {3 \choose 1} = 120$

c. $\dfrac{120}{495} \approx 0.242$


