---
title: "Homework 4"
author: "Mark Schulist"
format: pdf
pdf-engine: typst
---

## 3.3.3

a. 

$$
X = \text{amount spent on flat screen TVs by two random customers}
$$

```{=typst}
  #figure(
  table(
    stroke: black,
  	columns: (auto,auto),
   	[$x$],[$P(X = x)$],
   	[0],[`dbinom(0, 2, 0.3)` $= 0.49$],
   	[400],[$2 dot 0.7 dot 0.3 dot 0.4 = 0.168$],
   	[750],[$2 dot 0.7 dot 0.3 dot 0.6 = 0.252$],
   	[800],[`dbinom(2, 2, 0.3)` $dot 0.4^2 = 0.0144$],
   	[1150],[`dbinom(1, 2, 0.4)` $dot 0.3^2 = 0.0432$],
   	[1500],[`dbinom(2, 2, 0.3)` $dot 0.6^2 = 0.0324$]
  ),
  caption:[PMF of X]
  )
```

b. 

$$
E(X) = 400 \cdot 0.168 + 750 \cdot 0.252 + 800 \cdot 0.0144 + 1150 \cdot 0.0432 + 1500 \cdot 0.0324 = 366
$$

```{=typst}
$
"var"(X) = 0.49(366)^2 + 0.168(400 - 366)^2 + 0.252(750 - 366)^2 + \ 0.0144(800 - 366)^2 + 0.0432(1150 - 366)^2 + 0.0324(1500 - 366)^2 = 173922
$
```


## 3.3.7

a. 

```{=typst}
$
F(x) = x^2 / 4 
$

Median is when CDF = 0.5

$
0.5 &= x^2 / 4 \
2 &= x^2 \
sqrt(2) = x
$

Median is $sqrt(2)$. We can ignore negative case because $F(x)$ is only valid $0<x<=2$.

We can do the same for the 0.25 and 0.75 quantiles. 

$
0.25 &= x^2 / 4 \
1 &= x
$

$Q_1 = 1$

$
0.75 &= x^2 / 4 \
sqrt(3) &= x
$

$Q_3 = sqrt(3)$

IQR = $Q_3 - Q_1 = 0.732$
```


b.

```{=typst}
$
E(X) &= integral_0^2 x / 2 dot x d x \
&= lr(x^3 / 6 |)_(x=0)^(2) \
&= 4/3
$

$
"var"(X) &= E(X^2) - E(X)^2 \
&= integral_0^2 x / 2 dot x^2 d x - 16 / 9\
&= lr(x^4 / 8|)_(x=0)^(x=2) - 16 / 9\
&= 2 - 16 / 9 \
&= 2 / 9
$

"sd"(X) = sqrt(2/9) = 0.471
```


## 3.4.4

a.

```{=typst}
$
n = 10, p = 0.5
$

$
E(X) = n p = 5
$

$
"var"(X) = n p(1-p) = 2.5
$
```

b.

```{=typst}
$
P(X = 5) = binom(10, 5) 0.5^5 dot 0.5^5 = 0.246
$
```

c.

```{r}
most_5 <- pbinom(5, 10, 0.5)
```

The $P(X \leq 5) =$ `r most_5`.

d.

When $Y = 10 - X$, $Y$ represents the number of questions incorrectly answered. There are 10 total questions, so the complement $(10 - X)$ is the number that are not included in $X$. 

e.

We know the distribution is symmetric, so $P(2 \leq Y \leq 5) = P(2 \leq X \leq 5)$. 

```{r}
sym <- pbinom(5, 10, 0.5) - pbinom(1, 10, 0.5)
actual <- pbinom(8, 10, 0.5) - pbinom(4, 10, 0.5)
sym == actual
```

We can see that the two methods are equivalent, and both return a probability of `r sym`.

## 3.4.13

b.

$$
S = {0, 1, 2, 3}
$$

$$
\text{PMF}(X = x) = \frac{{3 \choose x}{17 \choose 5 - x}}{20 \choose 5}
$$

c.

$$
P(X = 1) = \frac{{3 \choose 1}{17 \choose 5 - 1}}{20 \choose 5} \approx 0.461
$$

d. 

```{=typst}
$
E(X) = 5 dot 3 / 20 = 3 / 4
$

$
"var"(X) = 3 / 4 (1 - 3 / 20) 15 / 19 approx 0.503
$

```

## Additional Problem

```{=typst}
$
f(x) = cases(
  1 "if" 0 <= x <= 1,
  0 "elsewhere"
)
$

$
"MGF"_(X)(t) &= E(e^(t x)) \
&= integral_0^1 e^(t x) 1 d x \
&= lr(e^(t x) / t|)_(x=0)^1 \
&= e^t / t - 1 / t \
&= (e^t - 1) / t
$
```

When $t \neq 0$, the moment generating function of $X$ is $\frac{e^t - 1}{t}$.

When $t = 0$, the MGF is equal to one. This is because the MGF when $t = 0$ takes the form of $E(X^0) = E(1) = 1$. 
